---
title: "Wage Pressure Tracker"
subtitle: "Tracking Wage and Labor Cost Pressures via Search Big Data Across Europe"
format:
  dashboard:
    theme:
      - flatly
      - theme.scss
    css: styles.css
    include-after-body: export_script.html
    orientation: rows
    scrolling: true
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(plotly)
library(dplyr)
library(tidyr)
library(readr)
library(htmltools)
library(DT)
library(jsonlite)
library(bigrquery)
library(lubridate)

# BigQuery Configuration
options(gargle_oauth_email = TRUE)
project <- "macroeconomic-dashboards"
dataset <- "Disinflation_dashboard"

# Looker Studio Core Color Palette
col_pink     <- "#d6498b"
col_yellow   <- "#fec44f"
col_orange   <- "#fe9929"
col_seledyn  <- "#41b6c4"
col_green    <- "#2ca25f"
col_purple   <- "#6a1b9a"
col_line     <- "#111111"

# Country codes, names, and distinct Looker Studio colors
country_codes <- c("PL", "DE", "ES", "NL", "FR", "IT")
country_names <- c("Poland", "Germany", "Spain", "Netherlands", "France", "Italy")
country_flags <- c("PL" = "🇵🇱", "DE" = "🇩🇪", "ES" = "🇪🇸", "NL" = "🇳🇱", "FR" = "🇫🇷", "IT" = "🇮🇹")

country_colors <- c(
  "PL" = "#d6498b", # Magenta / Pink
  "DE" = "#fec44f", # Warm Amber / Yellow
  "ES" = "#fe9929", # Orange
  "NL" = "#41b6c4", # Teal / Cyan
  "FR" = "#2ca25f", # Green
  "IT" = "#6a1b9a"  # Purple / Indigo
)

palette_components <- c("#d6498b", "#fec44f", "#fe9929", "#41b6c4")

# Data fetcher: tries Google BigQuery first, falls back to local precomputed CSVs
fetch_wage_pressure_data <- function() {
  tryCatch({
    bq_auth(email = TRUE)
    
    # 1. PCA Indices
    sql_pca <- paste0("SELECT date, geo, country, pca_raw, wage_pressure_index, wage_pressure_ma3m, wage_pressure_ma12m FROM `", 
                      project, ".", dataset, ".wage_spiral_pca_indices`")
    tb_pca <- bq_project_query(project, sql_pca)
    df_indices <- bq_table_download(tb_pca) %>%
      mutate(date = as.Date(date)) %>%
      arrange(geo, date)
      
    # 2. Components
    sql_comp <- paste0("SELECT date, country, geo, term, interest FROM `", 
                       project, ".", dataset, ".wage_spiral_tracker_long`")
    tb_comp <- bq_project_query(project, sql_comp)
    df_comp <- bq_table_download(tb_comp) %>%
      rename(Date = date, Country = country, Geo = geo, Term = term, Interest = interest) %>%
      mutate(Date = as.Date(Date)) %>%
      arrange(Geo, Term, Date)

    # 3. Loadings from local fallback (or derived)
    df_load <- if (file.exists("wage_spiral_pca_loadings.csv")) {
      read_csv("wage_spiral_pca_loadings.csv", show_col_types = FALSE)
    } else {
      data.frame()
    }
    
    # Update local caches
    tryCatch({
      write_csv(df_indices, "wage_spiral_pca_indices.csv")
      write_csv(df_comp, "wage_spiral_tracker_long.csv")
    }, error = function(e) {})
    
    list(indices = df_indices, components = df_comp, loadings = df_load)
  }, error = function(e) {
    message("Notice: BigQuery connection skipped (", e$message, "). Using local verified cache.")
    df_indices <- read_csv("wage_spiral_pca_indices.csv", show_col_types = FALSE) %>%
      mutate(date = as.Date(date))
    df_comp <- read_csv("wage_spiral_tracker_long.csv", show_col_types = FALSE) %>%
      mutate(Date = as.Date(Date))
    df_load <- read_csv("wage_spiral_pca_loadings.csv", show_col_types = FALSE)
    list(indices = df_indices, components = df_comp, loadings = df_load)
  })
}

data_pack <- fetch_wage_pressure_data()
df_ind   <- data_pack$indices
df_comp  <- data_pack$components
df_load  <- data_pack$loadings

# Load or calculate country statistics
stats_json_path <- "country_latest_stats.json"
if (file.exists(stats_json_path)) {
  stats_data <- jsonlite::fromJSON(stats_json_path)
} else {
  stats_data <- list()
}

# Precompute PCA loadings JSON for dynamic Plotly restyle (eliminates empty columns across countries)
pca_loadings_list <- list()
for (g in country_codes) {
  df_sub_load <- df_load %>% filter(geo == g) %>% arrange(desc(pca_loading))
  c_name <- country_names[which(country_codes == g)]
  terms_wrapped <- gsub(" \\+ ", "<br>+ ", df_sub_load$term)
  c_col <- unname(country_colors[g])
  colors <- rep(c_col, nrow(df_sub_load))
  hover <- paste0("<b>Query:</b> ", df_sub_load$term,
                  "<br><b>PCA Loading:</b> ", sprintf("%+.3f", df_sub_load$pca_loading),
                  "<br><b>Variance Explained:</b> ", df_sub_load$var_explained_pct, "%")
  
  pca_loadings_list[[g]] <- list(
    geo = g,
    country = c_name,
    terms = df_sub_load$term,
    terms_display = terms_wrapped,
    loadings = round(as.numeric(df_sub_load$pca_loading), 4),
    var_explained = as.numeric(unique(df_sub_load$var_explained_pct)[1]),
    colors = colors,
    hover = hover
  )
}
pca_loadings_json <- jsonlite::toJSON(pca_loadings_list, auto_unbox = TRUE)

# Helper function to generate clean 4-column factor loading charts by country
create_country_loading_plot <- function(geo_code) {
  df_sub <- df_load %>% 
    filter(geo == geo_code) %>% 
    arrange(desc(pca_loading)) %>%
    mutate(term_wrapped = gsub(" \\+ ", "<br>+ ", term))
  
  c_name <- country_names[which(country_codes == geo_code)]
  var_exp <- df_sub$var_explained_pct[1]
  c_col <- unname(country_colors[geo_code])
  bar_colors <- rep(c_col, nrow(df_sub))
  
  plot_ly(
    data = df_sub,
    x = ~term_wrapped,
    y = ~pca_loading,
    type = 'bar',
    width = 0.42,
    marker = list(
      color = bar_colors,
      line = list(color = "#0f172a", width = 1)
    ),
    showlegend = FALSE,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, " (", geo_code, "):</b> ", term,
                  "<br><b>PCA Loading:</b> ", sprintf("%+.3f", pca_loading),
                  "<br><b>Explained Variance:</b> ", var_exp, "%"),
    texttemplate = '%{y:+.2f}',
    textposition = 'outside',
    textfont = list(size = 12.5, color = "#0f172a", face = "bold")
  ) %>% layout(
    xaxis = list(
      title = "",
      categoryorder = 'array',
      categoryarray = df_sub$term_wrapped,
      tickfont = list(size = 11, color = "#1e293b")
    ),
    yaxis = list(
      title = list(text = "Factor Loading (PC1)", font = list(size = 12.5)),
      range = c(-0.45, 0.85),
      showgrid = TRUE,
      gridcolor = "#f0f0f0",
      zeroline = TRUE,
      zerolinecolor = "#64748b",
      zerolinewidth = 2
    ),
    margin = list(l = 50, r = 20, t = 30, b = 75, autoexpand = FALSE)
  ) %>%
  config(responsive = TRUE, displayModeBar = FALSE)
}

latest_date_str <- format(max(df_ind$date, na.rm = TRUE), "%B %Y")
earliest_date_str <- format(min(df_ind$date, na.rm = TRUE), "%Y")
```

```{=html}
<script>
  window.countryStatsData = `r if (file.exists("country_latest_stats.json")) paste(readLines("country_latest_stats.json", warn=FALSE), collapse="\n") else "{}"`;
  window.pcaLoadingsData = `r pca_loadings_json`;
</script>
```

# European Overview {icon="globe-europe"}

## Row {height=600px}

### Wage Pressure Across Major European Economies (Index) {width=100%}

::: {.chart-subtitle style="margin-bottom: 6px; color: #555555; font-size: 0.92rem;"}
Synthetic PCA index (0–100) based on Google Trends search queries, seasonally adjusted via STL
:::

```{r chart_cross_country_comparison}
# Main Cross-Country Comparison Chart (Fulfilling Looker Studio PDF Page 1)
p_main <- plot_ly()

# 1. 12-Month Moving Average Trend (Default - Active)
for (g in country_codes) {
  df_g <- df_ind %>% filter(geo == g) %>% arrange(date)
  c_name <- country_names[which(country_codes == g)]
  col <- country_colors[g]
  
  p_main <- p_main %>% add_trace(
    data = df_g, x = ~date, y = ~wage_pressure_ma12m,
    type = 'scatter', mode = 'lines',
    name = c_name,
    line = list(color = col, width = 2.6),
    visible = TRUE,
    legendgroup = g,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, " (", g, ")</b>",
                  "<br><b>Period:</b> ", format(date, "%B %Y"),
                  "<br><b>Wage Pressure (MA 12M):</b> ", sprintf("%.1f", wage_pressure_ma12m), " / 100",
                  "<br><b>Short-term (MA 3M):</b> ", sprintf("%.1f", wage_pressure_ma3m),
                  "<br><b>Monthly SA:</b> ", sprintf("%.1f", wage_pressure_index))
  )
}

# 2. 3-Month Moving Average Momentum (Visible via switcher)
for (g in country_codes) {
  df_g <- df_ind %>% filter(geo == g) %>% arrange(date)
  c_name <- country_names[which(country_codes == g)]
  col <- country_colors[g]
  
  p_main <- p_main %>% add_trace(
    data = df_g, x = ~date, y = ~wage_pressure_ma3m,
    type = 'scatter', mode = 'lines',
    name = paste0(c_name, " (MA 3M)"),
    line = list(color = col, width = 2.0, dash = 'dot'),
    visible = FALSE,
    legendgroup = g,
    showlegend = FALSE,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, " (MA 3M)</b>",
                  "<br><b>Period:</b> ", format(date, "%B %Y"),
                  "<br><b>Momentum (MA 3M):</b> ", sprintf("%.1f", wage_pressure_ma3m), " / 100")
  )
}

# 3. Monthly DeseSeasonalized Index (Visible via switcher)
for (g in country_codes) {
  df_g <- df_ind %>% filter(geo == g) %>% arrange(date)
  c_name <- country_names[which(country_codes == g)]
  col <- country_colors[g]
  
  p_main <- p_main %>% add_trace(
    data = df_g, x = ~date, y = ~wage_pressure_index,
    type = 'scatter', mode = 'lines',
    name = paste0(c_name, " (Monthly SA)"),
    line = list(color = col, width = 1.2, dash = 'dash'),
    visible = FALSE,
    legendgroup = g,
    showlegend = FALSE,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, " (Monthly SA)</b>",
                  "<br><b>Period:</b> ", format(date, "%B %Y"),
                  "<br><b>Monthly SA Index:</b> ", sprintf("%.1f", wage_pressure_index), " / 100")
  )
}

n_countries <- length(country_codes)
vis_ma12 <- c(rep(TRUE, n_countries),  rep(FALSE, n_countries), rep(FALSE, n_countries))
vis_ma3  <- c(rep(FALSE, n_countries), rep(TRUE, n_countries),  rep(FALSE, n_countries))
vis_raw  <- c(rep(FALSE, n_countries), rep(FALSE, n_countries), rep(TRUE, n_countries))

p_main <- p_main %>% layout(
  xaxis = list(
    title = "",
    dtick = "M24",
    tickformat = "%Y",
    showgrid = FALSE,
    tickfont = list(size = 13, color = "#222222"),
    rangeselector = list(
      buttons = list(
        list(count = 3, label = "3Y", step = "year", stepmode = "backward"),
        list(count = 5, label = "5Y", step = "year", stepmode = "backward"),
        list(count = 10, label = "10Y", step = "year", stepmode = "backward"),
        list(step = "all", label = "All Time")
      ),
      x = 0.70, y = 1.15,
      yanchor = "bottom",
      bgcolor = "#f1f5f9",
      activecolor = "#fec44f"
    )
  ),
  yaxis = list(
    title = list(text = "Wage Pressure Index (0–100)", font = list(size = 14, color = "#111111")),
    range = c(-2, 105),
    showgrid = TRUE,
    gridcolor = "#f0f0f0",
    tickfont = list(size = 13, color = "#222222")
  ),
  margin = list(l = 50, r = 25, t = 80, b = 40, autoexpand = FALSE),
  legend = list(
    orientation = "h",
    x = 0.01, y = 1.01,
    yanchor = "bottom",
    font = list(size = 12.5, color = "#1e293b"),
    bgcolor = "rgba(255, 255, 255, 0.9)"
  ),
  updatemenus = list(
    list(
      type = "dropdown",
      active = 0,
      x = 0.01, y = 1.15,
      xanchor = "left", yanchor = "bottom",
      direction = "down",
      pad = list(r = 10, t = 2, b = 2),
      buttons = list(
        list(method = "restyle", args = list("visible", vis_ma12), label = "Smoothing: 12-Month Trend (MA 12M)"),
        list(method = "restyle", args = list("visible", vis_ma3),  label = "Smoothing: 3-Month Momentum (MA 3M)"),
        list(method = "restyle", args = list("visible", vis_raw),  label = "Smoothing: Monthly SA Index (Raw)")
      )
    )
  ),
  hovermode = 'x unified',
  hoverlabel = list(font = list(size = 13))
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_main
```

## Row {height=420px}

### Cross-Country Wage Pressure Benchmarking Table {width=100%}

```{r table_cross_country}
summary_table_data <- data.frame(
  Flag = sapply(country_codes, function(g) country_flags[g]),
  Country = country_names,
  Code = country_codes,
  Current_Level = sapply(country_codes, function(g) stats_data[[g]]$current_ma12),
  YoY_Change = sapply(country_codes, function(g) stats_data[[g]]$yoy_change),
  Change_3Y = sapply(country_codes, function(g) stats_data[[g]]$change_3y),
  Historical_Peak = sapply(country_codes, function(g) stats_data[[g]]$peak_value),
  Peak_Period = sapply(country_codes, function(g) stats_data[[g]]$peak_period),
  Explained_Variance = sapply(country_codes, function(g) paste0(stats_data[[g]]$var_explained, "%")),
  Status = sapply(country_codes, function(g) stats_data[[g]]$status),
  stringsAsFactors = FALSE
) %>%
arrange(desc(Current_Level))

datatable(
  summary_table_data,
  rownames = FALSE,
  colnames = c(" ", "Country", "ISO", "Latest Level (0–100)", "1-Year YoY (pts)", "3-Year Trend (pts)", "All-Time Peak", "Peak Date", "PCA Variance Explained", "Pressure Status"),
  options = list(
    pageLength = 6,
    dom = 't',
    ordering = TRUE,
    paging = FALSE,
    searching = FALSE,
    info = FALSE,
    autoWidth = FALSE,
    columnDefs = list(
      list(className = 'dt-center', targets = c(0, 2, 7, 8, 9)),
      list(className = 'dt-right', targets = c(3, 4, 5, 6))
    )
  )
) %>%
formatStyle(
  'Current_Level',
  background = styleColorBar(c(0, 100), '#fec44f'),
  backgroundSize = '95% 70%',
  backgroundRepeat = 'no-repeat',
  backgroundPosition = 'center'
) %>%
formatStyle(
  'YoY_Change',
  color = styleInterval(0, c('#059669', '#d97706')),
  fontWeight = 'bold'
)
```

# Germany & France {icon="diagram-2"}

## Row {height=520px}

### Germany - Components Breakdown {width=100%}

```{r chart_germany_components}
# Germany Components (Corrected from draft Looker Studio slide)
df_de <- df_comp %>% filter(Geo == "DE") %>% arrange(Date)
terms_de <- unique(df_de$Term)

p_de <- plot_ly()
for (i in seq_along(terms_de)) {
  t <- terms_de[i]
  sub_t <- df_de %>% filter(Term == t)
  col <- palette_components[(i - 1) %% length(palette_components) + 1]
  
  p_de <- p_de %>% add_trace(
    data = sub_t, x = ~Date, y = ~Interest,
    type = 'scatter', mode = 'lines',
    name = t,
    line = list(color = col, width = 2.2),
    hoverinfo = 'text',
    text = ~paste0("<b>Germany: ", t, "</b>",
                  "<br><b>Period:</b> ", format(Date, "%B %Y"),
                  "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
  )
}

p_de <- p_de %>% layout(
  title = list(
    text = "<b>Germany — Deseasonalized Components (Google Trends SA Index 0–100)</b><br><span style='font-size:13px;color:#555555;'>Underlying collective bargaining, salary negotiation, and minimum wage search terms</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_de
```

## Row {height=520px}

### France - Components Breakdown {width=100%}

```{r chart_france_components}
df_fr <- df_comp %>% filter(Geo == "FR") %>% arrange(Date)
terms_fr <- unique(df_fr$Term)

p_fr <- plot_ly()
for (i in seq_along(terms_fr)) {
  t <- terms_fr[i]
  sub_t <- df_fr %>% filter(Term == t)
  col <- palette_components[(i - 1) %% length(palette_components) + 1]
  
  p_fr <- p_fr %>% add_trace(
    data = sub_t, x = ~Date, y = ~Interest,
    type = 'scatter', mode = 'lines',
    name = t,
    line = list(color = col, width = 2.2),
    hoverinfo = 'text',
    text = ~paste0("<b>France: ", t, "</b>",
                  "<br><b>Period:</b> ", format(Date, "%B %Y"),
                  "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
  )
}

p_fr <- p_fr %>% layout(
  title = list(
    text = "<b>France — Deseasonalized Components (Google Trends SA Index 0–100)</b><br><span style='font-size:13px;color:#555555;'>Underlying NAO, SMIC, and salary revaluation search queries</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_fr
```

# Netherlands & Italy {icon="diagram-3"}

## Row {height=520px}

### Netherlands - Components Breakdown {width=100%}

```{r chart_netherlands_components}
df_nl <- df_comp %>% filter(Geo == "NL") %>% arrange(Date)
terms_nl <- unique(df_nl$Term)

p_nl <- plot_ly()
for (i in seq_along(terms_nl)) {
  t <- terms_nl[i]
  sub_t <- df_nl %>% filter(Term == t)
  col <- palette_components[(i - 1) %% length(palette_components) + 1]
  
  p_nl <- p_nl %>% add_trace(
    data = sub_t, x = ~Date, y = ~Interest,
    type = 'scatter', mode = 'lines',
    name = t,
    line = list(color = col, width = 2.2),
    hoverinfo = 'text',
    text = ~paste0("<b>Netherlands: ", t, "</b>",
                  "<br><b>Period:</b> ", format(Date, "%B %Y"),
                  "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
  )
}

p_nl <- p_nl %>% layout(
  title = list(
    text = "<b>Netherlands — Deseasonalized Components (Google Trends SA Index 0–100)</b><br><span style='font-size:13px;color:#555555;'>Underlying CAO negotiations, loonstijging, and minimumloon search terms</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_nl
```

## Row {height=520px}

### Italy - Components Breakdown {width=100%}

```{r chart_italy_components}
df_it <- df_comp %>% filter(Geo == "IT") %>% arrange(Date)
terms_it <- unique(df_it$Term)

p_it <- plot_ly()
for (i in seq_along(terms_it)) {
  t <- terms_it[i]
  sub_t <- df_it %>% filter(Term == t)
  col <- palette_components[(i - 1) %% length(palette_components) + 1]
  
  p_it <- p_it %>% add_trace(
    data = sub_t, x = ~Date, y = ~Interest,
    type = 'scatter', mode = 'lines',
    name = t,
    line = list(color = col, width = 2.2),
    hoverinfo = 'text',
    text = ~paste0("<b>Italy: ", t, "</b>",
                  "<br><b>Period:</b> ", format(Date, "%B %Y"),
                  "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
  )
}

p_it <- p_it %>% layout(
  title = list(
    text = "<b>Italy — Deseasonalized Components (Google Trends SA Index 0–100)</b><br><span style='font-size:13px;color:#555555;'>Underlying CCNL renewals, aumento stipendio, and adeguamento salariale queries</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_it
```

# Spain & Poland {icon="diagram-3-fill"}

## Row {height=520px}

### Spain - Components Breakdown {width=100%}

```{r chart_spain_components}
df_es <- df_comp %>% filter(Geo == "ES") %>% arrange(Date)
terms_es <- unique(df_es$Term)

p_es <- plot_ly()
for (i in seq_along(terms_es)) {
  t <- terms_es[i]
  sub_t <- df_es %>% filter(Term == t)
  col <- palette_components[(i - 1) %% length(palette_components) + 1]
  
  p_es <- p_es %>% add_trace(
    data = sub_t, x = ~Date, y = ~Interest,
    type = 'scatter', mode = 'lines',
    name = t,
    line = list(color = col, width = 2.2),
    hoverinfo = 'text',
    text = ~paste0("<b>Spain: ", t, "</b>",
                  "<br><b>Period:</b> ", format(Date, "%B %Y"),
                  "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
  )
}

p_es <- p_es %>% layout(
  title = list(
    text = "<b>Spain — Deseasonalized Components (Google Trends SA Index 0–100)</b><br><span style='font-size:13px;color:#555555;'>Underlying convenio colectivo, subida SMI, and revisión salarial queries</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_es
```

## Row {height=520px}

### Poland - Components Breakdown {width=100%}

```{r chart_poland_components}
df_pl <- df_comp %>% filter(Geo == "PL") %>% arrange(Date)
terms_pl <- unique(df_pl$Term)

p_pl <- plot_ly()
for (i in seq_along(terms_pl)) {
  t <- terms_pl[i]
  sub_t <- df_pl %>% filter(Term == t)
  col <- palette_components[(i - 1) %% length(palette_components) + 1]
  
  p_pl <- p_pl %>% add_trace(
    data = sub_t, x = ~Date, y = ~Interest,
    type = 'scatter', mode = 'lines',
    name = t,
    line = list(color = col, width = 2.2),
    hoverinfo = 'text',
    text = ~paste0("<b>Poland: ", t, "</b>",
                  "<br><b>Period:</b> ", format(Date, "%B %Y"),
                  "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
  )
}

p_pl <- p_pl %>% layout(
  title = list(
    text = "<b>Poland — Deseasonalized Components (Google Trends SA Index 0–100)</b><br><span style='font-size:13px;color:#555555;'>Underlying podwyżka płacy, negocjacja pensji, and kalkulator wynagrodzeń queries</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_pl
```

# Country Explorer {icon="search"}

## Row {height=130px}

```{r country_explorer_controls}
# Native Country Selector Component
tags$div(
  class = "country-filter-card",
  tags$div(
    class = "country-filter-content",
    tags$label(
      class = "country-filter-label",
      `for` = "select-country-explorer",
      "Select an economy to explore its synthetic wage index, components, and PCA structure:"
    ),
    tags$div(
      class = "country-select-box-wrapper",
      tags$span(class = "country-select-prefix", "Economy:"),
      tags$select(
        id = "select-country-explorer",
        class = "form-select country-native-select",
        onchange = "window.onCountryExplorerChange(this.value)",
        lapply(seq_along(country_codes), function(idx) {
          code <- country_codes[idx]
          name <- country_names[idx]
          sel_attr <- if (code == "PL") "selected" else NULL
          tags$option(value = code, selected = sel_attr, paste0(country_flags[code], " ", code, " - ", name))
        })
      )
    )
  )
)
```

## Row {height=520px}

### Synthetic Wage Pressure Index for Selected Economy {width=100%}

```{r explorer_plot_index}
# Multi-trace plot with all countries, toggled by JS
p_exp_idx <- plot_ly()

for (i in seq_along(country_codes)) {
  g <- country_codes[i]
  c_name <- country_names[i]
  df_g <- df_ind %>% filter(geo == g) %>% arrange(date)
  is_active <- (i == 1) # PL active by default
  col_theme <- country_colors[g]
  
  # Trace 1: Raw monthly index
  p_exp_idx <- p_exp_idx %>% add_trace(
    data = df_g, x = ~date, y = ~wage_pressure_index,
    type = 'scatter', mode = 'lines',
    name = paste0(c_name, ' (Monthly SA)'),
    line = list(color = '#94a3b8', width = 1.0, dash = 'solid'),
    opacity = 0.55,
    visible = is_active,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, "</b>",
                  "<br><b>Date:</b> ", format(date, "%B %Y"),
                  "<br><b>Monthly SA Index:</b> ", sprintf("%.1f", wage_pressure_index))
  )
  
  # Trace 2: MA 3M momentum
  p_exp_idx <- p_exp_idx %>% add_trace(
    data = df_g, x = ~date, y = ~wage_pressure_ma3m,
    type = 'scatter', mode = 'lines',
    name = paste0(c_name, ' (MA 3M)'),
    line = list(color = '#0284c7', width = 1.8),
    visible = is_active,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, "</b>",
                  "<br><b>Date:</b> ", format(date, "%B %Y"),
                  "<br><b>Short-term (MA 3M):</b> ", sprintf("%.1f", wage_pressure_ma3m))
  )
  
  # Trace 3: MA 12M trend with shaded area
  p_exp_idx <- p_exp_idx %>% add_trace(
    data = df_g, x = ~date, y = ~wage_pressure_ma12m,
    type = 'scatter', mode = 'lines',
    name = paste0(c_name, ' (MA 12M Trend)'),
    line = list(color = col_theme, width = 2.8),
    fill = 'tozeroy', fillcolor = paste0(col_theme, "22"),
    visible = is_active,
    hoverinfo = 'text',
    text = ~paste0("<b>", c_name, "</b>",
                  "<br><b>Date:</b> ", format(date, "%B %Y"),
                  "<br><b>Wage Pressure Trend (MA 12M):</b> ", sprintf("%.1f", wage_pressure_ma12m))
  )
}

p_exp_idx <- p_exp_idx %>% layout(
  title = list(
    text = "<b>Synthetic Wage Pressure Index: Monthly SA vs Moving Averages</b><br><span style='font-size:13px;color:#555555;'>Monthly deseasonalized index, 3-month momentum, and 12-month structural trend (0–100 scale)</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Index (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 12)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_exp_idx
```

## Row {height=500px}

### Component Dynamics for Selected Economy {width=100%}

```{r explorer_plot_components}
p_exp_comp <- plot_ly()

for (i in seq_along(country_codes)) {
  g <- country_codes[i]
  c_name <- country_names[i]
  df_sub <- df_comp %>% filter(Geo == g) %>% arrange(Date)
  terms <- unique(df_sub$Term)
  is_active <- (i == 1)
  
  for (j in seq_along(terms)) {
    t <- terms[j]
    sub_t <- df_sub %>% filter(Term == t)
    col <- palette_components[(j - 1) %% length(palette_components) + 1]
    
    p_exp_comp <- p_exp_comp %>% add_trace(
      data = sub_t, x = ~Date, y = ~Interest,
      type = 'scatter', mode = 'lines',
      name = t,
      line = list(color = col, width = 2.2),
      visible = is_active,
      hoverinfo = 'text',
      text = ~paste0("<b>", t, "</b>",
                    "<br><b>Period:</b> ", format(Date, "%B %Y"),
                    "<br><b>Intensity (SA 0–100):</b> ", sprintf("%.1f", Interest))
    )
  }
}

p_exp_comp <- p_exp_comp %>% layout(
  title = list(
    text = "<b>Underlying Google Trends Search Components (SA 0–100)</b><br><span style='font-size:13px;color:#555555;'>Comparative time series of each individual search query pillar</span>",
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.18,
    yanchor = "top"
  ),
  xaxis = list(title = "", dtick = "M24", tickformat = "%Y", showgrid = FALSE, tickfont = list(size = 13, color = "#222222")),
  yaxis = list(title = list(text = "Normalized Intensity (0–100)", font = list(size = 13.5)), range = c(-2, 105), showgrid = TRUE, gridcolor = "#f0f0f0"),
  margin = list(l = 50, r = 25, t = 115, b = 40, autoexpand = FALSE),
  legend = list(orientation = "h", x = 0.01, y = 1.02, yanchor = "bottom", font = list(size = 11.5)),
  hovermode = 'x unified'
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_exp_comp
```

## Row {height=450px}

### PCA Factor Loadings (Component Weights) {width=100%}

```{r explorer_plot_loadings}
# Initial country is PL (Poland) - exactly 1 trace to avoid empty column slots
init_geo <- "PL"
df_init_load <- df_load %>% filter(geo == init_geo) %>% arrange(desc(pca_loading)) %>%
  mutate(term_wrapped = gsub(" \\+ ", "<br>+ ", term))

c_col_init <- unname(country_colors[init_geo])
bar_colors_init <- rep(c_col_init, nrow(df_init_load))

p_exp_load <- plot_ly(
  data = df_init_load,
  x = ~term_wrapped, y = ~pca_loading,
  type = 'bar',
  width = 0.45,
  marker = list(
    color = bar_colors_init,
    line = list(color = "#0f172a", width = 1)
  ),
  showlegend = FALSE,
  hoverinfo = 'text',
  text = ~paste0("<b>Query:</b> ", term,
                "<br><b>PCA Loading:</b> ", sprintf("%+.3f", pca_loading),
                "<br><b>Variance Explained:</b> ", var_explained_pct, "%"),
  texttemplate = '%{y:+.2f}',
  textposition = 'outside',
  textfont = list(size = 13, color = "#0f172a", face = "bold")
) %>% layout(
  title = list(
    text = paste0("<b>PCA Factor Loadings — Poland (PC1 Weights)</b><br><span style='font-size:13px;color:#555555;'>Contribution of each search phrase to PC1 (", df_init_load$var_explained_pct[1], "% variance explained)</span>"),
    font = list(size = 15, color = "#111111"),
    x = 0.01, y = 1.16,
    yanchor = "top"
  ),
  xaxis = list(
    title = "",
    categoryorder = 'array',
    categoryarray = df_init_load$term_wrapped,
    tickfont = list(size = 11.5, color = "#1e293b")
  ),
  yaxis = list(
    title = list(text = "Factor Loading (Weight in PC1)", font = list(size = 13.5)),
    range = c(-0.45, 0.85),
    showgrid = TRUE,
    gridcolor = "#f0f0f0",
    zeroline = TRUE,
    zerolinecolor = "#64748b",
    zerolinewidth = 2
  ),
  margin = list(l = 55, r = 25, t = 95, b = 90, autoexpand = FALSE)
) %>%
config(responsive = TRUE, displayModeBar = FALSE)

p_exp_load
```

# Methodology & PCA Weights {icon="info-circle"}

## Row {height=420px}

### 🇵🇱 Poland — Factor Loadings (88.2% Variance Explained) {width=50%}

```{r chart_loadings_pl}
create_country_loading_plot("PL")
```

### 🇩🇪 Germany — Factor Loadings (60.8% Variance Explained) {width=50%}

```{r chart_loadings_de}
create_country_loading_plot("DE")
```

## Row {height=420px}

### 🇪🇸 Spain — Factor Loadings (71.7% Variance Explained) {width=50%}

```{r chart_loadings_es}
create_country_loading_plot("ES")
```

### 🇳🇱 Netherlands — Factor Loadings (93.4% Variance Explained) {width=50%}

```{r chart_loadings_nl}
create_country_loading_plot("NL")
```

## Row {height=420px}

### 🇫🇷 France — Factor Loadings (72.2% Variance Explained) {width=50%}

```{r chart_loadings_fr}
create_country_loading_plot("FR")
```

### 🇮🇹 Italy — Factor Loadings (71.3% Variance Explained) {width=50%}

```{r chart_loadings_it}
create_country_loading_plot("IT")
```

## Row {width=100%}

```{=html}
<div class="methodology-header">
  <span>💡 <strong>Card 1: Motivation &amp; Economic Intuition — Google Trends Big Data as a Leading Wage Indicator</strong></span>
  <span class="badge-tag">Motivation &amp; 4 Pillars</span>
</div>
```

::: {.card-body .methodology-card}
#### **Why Track Wage Pressures Using Search Query Big Data?**

1. **The Severe Publication Lag of National Accounts**:
   - Official compensation indicators (such as Eurostat's **Labor Cost Index (LCI)** and national accounts compensation per employee) are published with a lag of **90 to 120 days** after the close of a quarter.
   - By the time official statistics report an acceleration in unit labor costs, wage negotiations are already concluded and secondary price pass-through into service prices has materialized.
2. **Behavioral Footprints of Labor Dynamics**:
   - Google Trends query volumes provide an unprompted, high-frequency behavioral reflection of labor market tensions.
   - When inflation erodes real household purchasing power, searches for wage renegotiation and salary raises surge. Similarly, when trade unions prepare collective bargaining rounds (*Tarifverhandlungen*, *convenios colectivos*, *rinnovi CCNL*), public search activity peaks **months before** formal wage agreements take legal effect.
3. **Four Core Query Pillars Tracked per Economy**:
   - **Collective Bargaining**: Searches monitoring union wage negotiations and national sector settlements.
   - **Direct Salary Negotiation**: Searches related to requesting raises, asking for higher pay, and salary increase letters.
   - **Statutory Minimum Wage**: Searches regarding government minimum wage revisions and mandatory revaluations.
   - **Earnings & Take-Home Pay Calculators**: Queries evaluating net vs. gross compensation and purchasing power.
4. **Synthetic Leading Indicator**:
   - Extracting the common latent driver across these four pillars creates a real-time, forward-looking barometer of structural wage pressure across European economies.
:::

## Row {width=100%}

```{=html}
<div class="methodology-header">
  <span>📐 <strong>Card 2: Econometric Pipeline — Loess STL, PCA Factor Extraction &amp; Index Normalization (0–100)</strong></span>
  <span class="badge-tag">Econometric Model</span>
</div>
```

::: {.card-body .methodology-card}
#### **Statistical Methodology and Latent Factor Extraction**

Let $Y_{i,c,t}$ represent the raw Google Trends search volume index for term $i \in \{1, \dots, p\}$ in country $c$ at month $t$.

1. **Sampling Denoising**: High-frequency sampling jitter is filtered using a 3-month center moving average:

$$\tilde{Y}_{i,c,t} = \frac{1}{4} Y_{i,c,t-1} + \frac{1}{2} Y_{i,c,t} + \frac{1}{4} Y_{i,c,t+1}$$

2. **Adaptive Seasonal Adjustment (STL)**: Decomposed into trend ($T$), seasonal ($S$), and remainder ($R$) components using Loess with seasonal window $s.\text{window} = 7$:

$$\tilde{Y}_{i,c,t} = T_{i,c,t} + S_{i,c,t} + R_{i,c,t} \implies Y^{\text{SA}}_{i,c,t} = \tilde{Y}_{i,c,t} - S_{i,c,t}$$

3. **Standardization (Z-Score)**: Each seasonally adjusted component is standardized to zero mean and unit variance:

$$Z_{i,c,t} = \frac{Y^{\text{SA}}_{i,c,t} - \mu_{i,c}}{\sigma_{i,c}}, \quad \mathbf{Z}_c \in \mathbb{R}^{T \times p}$$

4. **Principal Component Analysis (PCA)**: The empirical covariance matrix $\boldsymbol{\Sigma}_c = \frac{1}{T-1} \mathbf{Z}_c^T \mathbf{Z}_c$ is decomposed:

$$\boldsymbol{\Sigma}_c \mathbf{v}_1 = \lambda_1 \mathbf{v}_1$$

where $\lambda_1$ is the largest eigenvalue and $\mathbf{v}_1 = [v_{1,1}, \dots, v_{1,p}]^T$ is the first eigenvector (factor loadings). The variance explained ratio confirms the validity of a unified wage-pressure factor ($60.8\% - 93.4\%$ across EU countries):

$$\operatorname{VarExplained}_c = \frac{\lambda_1}{\sum_{j=1}^p \lambda_j} \times 100\%$$

5. **Factor Projection & Min-Max Index Normalization (0–100 Scale)**:

$$\operatorname{PC1}_{c,t} = \operatorname{sgn}\left(\sum_{j=1}^p v_{1,j}\right) \cdot \mathbf{Z}_{c,t} \mathbf{v}_1$$

$$I_{c,t} = 100 \times \left( \frac{\operatorname{PC1}_{c,t} - \min_\tau \operatorname{PC1}_{c,\tau}}{\max_\tau \operatorname{PC1}_{c,\tau} - \min_\tau \operatorname{PC1}_{c,\tau}} \right)$$

6. **Momentum (MA 3M) and Structural Trend (MA 12M)**:

$$\text{Momentum}_{c,t} = \frac{1}{3}\sum_{k=0}^2 I_{c,t-k}, \quad \text{Trend}_{c,t} = \frac{1}{12}\sum_{k=0}^{11} I_{c,t-k}$$
:::


